---
title: "Choosing a bias-correction method"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing a bias-correction method}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6, fig.height = 3.4, dpi = 96, dev = "png"
)
options(width = 90)
```

`fit_met_bias_correction()` offers five transfer functions. This vignette
compares them on the bundled Lake Rotorua buoy record and shows how to read
the leave-one-year-out (LOYO) skill table, so you can pick deliberately
rather than taking the default on faith.

```{r setup, message = FALSE, warning = FALSE}
library(metscale)
library(ggplot2)
theme_set(theme_bw(base_size = 10) +
          theme(panel.grid.minor = element_blank(), legend.position = "top",
                legend.title = element_blank()))

ex <- system.file("extdata", package = "metscale")
tz <- "Etc/GMT-12"

obs <- prepare_obs_met(file.path(ex, "rotorua_buoy_met_aeme_hr.csv.gz"),
                       resample = "hour", tz = tz, wind_height = 2,
                       station = "rotorua_buoy", verbose = FALSE)

era5 <- read.csv(file.path(ex, "rotorua_era5_hourly_met.csv.gz"), check.names = FALSE)
era5$Date <- as.POSIXct(era5$Date, tz = tz, format = "%Y-%m-%d %H:%M:%S")
```

## The five methods

| `method` | transfer function | shape it can change |
|---|---|---|
| `"scale"` | per-month additive offset or ratio | level only |
| `"variance"` | per-month mean **and** standard-deviation matching | level + spread |
| `"linear"` | per-month least squares `obs ~ a + b * era5` (`robust` optional) | level + linear slope |
| `"eqm"` | empirical quantile mapping: remap the ERA5 CDF onto the observed CDF | the whole distribution |
| `"qdm"` | quantile **delta** mapping: as `eqm` but keeps the ERA5 anomaly relative to its training quantile | distribution, trend-preserving |

`transform` (`"auto"` / `"additive"` / `"ratio"`) applies to `"scale"`;
`by` (`"doy-loess"` / `"month"` / `"none"`) applies to `"scale"` and is
coerced to `"month"` for the others. `cv = "loyo"` (default) adds the
out-of-sample rows to the skill table --- always look at those, not the
in-sample `corrected` rows.

## Fit and compare

```{r fits}
vars    <- c("MET_tmpair", "MET_wndspd", "MET_humrel")
methods <- c("scale", "variance", "eqm", "qdm")

fits <- lapply(methods, function(m)
  fit_met_bias_correction(era5, obs, vars = vars, method = m, verbose = FALSE))
names(fits) <- methods

skill <- do.call(rbind, Map(function(f, m)
  cbind(method = m, f$skill[f$skill$stage == "cv_corrected", ]),
  fits, names(fits)))
raw <- transform(fits[[1]]$skill[fits[[1]]$skill$stage == "cv_raw", ],
                 method = "raw ERA5")
skill <- rbind(raw[names(skill)], skill)
skill[, c("method", "variable", "bias", "mae", "rmse", "r", "kge")]
```

```{r skill-fig, fig.height = 3, fig.cap = "**Figure 1.** Leave-one-year-out RMSE by method, dashed line = raw ERA5, per variable. All four cut the error sharply, but `scale` --- which only shifts the monthly level --- trails `variance` / `eqm` / `qdm` for air temperature and humidity, because ERA5's error at this site is partly a *variance* error that a pure offset cannot reach."}
ref <- skill[skill$method == "raw ERA5", c("variable", "rmse")]

ggplot(subset(skill, method != "raw ERA5"),
       aes(method, rmse, fill = method)) +
  geom_col(width = 0.65) +
  geom_hline(data = ref, aes(yintercept = rmse), linetype = 2) +
  facet_wrap(~ variable, scales = "free_y") +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  labs(x = NULL, y = "cross-validated RMSE (dashed = raw ERA5)")
```

## Which to use

The choice is a trade-off between present-day fit and how the fit behaves
when you push it outside its training range.

**For a present-day correction** --- the corrected series just has to match
the observed climate over the same period --- use the most skilful method.
Here that is `variance`, `eqm` or `qdm`: they cut the LOYO air-temperature
RMSE from 2.0 &deg;C (`scale`) to about 1.4 &deg;C and lift its KGE from
0.78 to 0.94, because they correct the day-to-day spread and not only the
level. `eqm` / `qdm` additionally fix a skewed distribution (wind speed,
humidity).

**For a baseline that will be projected** --- a delta-change, or any
extrapolation beyond the training quantiles --- a fitted CDF (`eqm`, `qdm`)
has nothing to say outside its training range and extrapolates with a flat
tail, which distorts a warmed future series. `vignette("scenario-workflow")`
therefore uses `"scale"` for maximum caution. `variance` is a reasonable
middle ground when the present-day variance error is large, as it is here:
it is still a linear transform, so it extrapolates predictably, while
recovering most of the spread that `scale` leaves behind.

## `by = "month"` vs `"doy-loess"`

`"month"` fits 12 independent offsets; `"doy-loess"` smooths them across the
day-of-year so the correction does not jump at month boundaries. The
smoothed version is usually a little better out of sample and always
gentler on a continuous series:

```{r by}
b_month <- fit_met_bias_correction(era5, obs, vars = "MET_tmpair",
                                   method = "scale", by = "month", verbose = FALSE)
b_loess <- fit_met_bias_correction(era5, obs, vars = "MET_tmpair",
                                   method = "scale", by = "doy-loess", verbose = FALSE)

rbind(month = b_month$skill[b_month$skill$stage == "cv_corrected", ],
      loess = b_loess$skill[b_loess$skill$stage == "cv_corrected", ])[, c("bias", "rmse", "kge")]
```

## Mixing methods per variable

`method` (and `transform`) accept a named vector, so you can quantile-map
the skewed fields and keep a plain offset elsewhere:

```{r per-var}
bc <- fit_met_bias_correction(
  era5, obs,
  vars   = c("MET_tmpair", "MET_wndspd", "MET_humrel"),
  method = c(MET_tmpair = "scale", MET_wndspd = "qdm", MET_humrel = "qdm"),
  verbose = FALSE)
bc$skill[bc$skill$stage == "cv_corrected", c("variable", "bias", "rmse", "kge")]
```

Apply any of these with `apply_met_bias_correction()` (whole record) or
`bias_correct_daily_baseline()` (record &rarr; corrected daily baseline).

## See also

* `vignette("scenario-workflow")` --- why the projected baseline uses
  `"scale"`, and the delta-change / disaggregation steps that follow.
* `vignette("met-data-frames")` --- preparing the ERA5 and observation
  frames that go in.
