---
title: "Climate extremes as impact-model forcing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Climate extremes as impact-model forcing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6, fig.height = 3.4, dpi = 96, dev = "png"
)
options(width = 90)
```

A lake model responds to the *tails* of its meteorology, not only the
seasonal means. Stratification breaks down when a wind burst clears some
threshold; a full turnover needs a cold outbreak; bottom-water anoxia
follows a long warm calm spell; a nutrient flush follows a single
heavy-rain day. The 20-year mean climate can barely move while any of
these change substantially.

`vignette("scenario-workflow")` builds a projection by adding one
monthly-mean change factor per variable to an observation-anchored
baseline. That is the right default for bulk energy-balance behaviour, but
by construction it cannot represent either of the things that matter for
extremes:

1. **a change in the *shape* of the distribution** --- the 99th percentile
   of daily rainfall moving by more (or less) than the mean;
2. **a change in the *frequency, duration or timing* of events** --- more
   wet days, longer dry spells, a storm season that shifts.

This vignette shows three complementary ways to keep extremes in the
forcing, each leaving a different aspect of the future free to differ from
the past --- the distribution's shape, the frequency and timing of events,
or the intensity of one specific event --- then how to check the result
and hand it to a process model.

```{r setup, message = FALSE, warning = FALSE}
library(metscale)
library(ggplot2)

theme_set(
  theme_bw(base_size = 10) +
    theme(panel.grid.minor = element_blank(),
          legend.position  = "top",
          legend.title     = element_blank(),
          strip.background  = element_blank(),
          strip.placement   = "outside")
)

ex   <- system.file("extdata", package = "metscale")
tz   <- "Etc/GMT-12"
lon  <- 176.2717
lat  <- -38.0790
elev <- 279
```

The bundled data is the same small Lake Rotorua slice used elsewhere: a
three-year hourly ERA5-Land extract, the matching buoy record, and the
NIWA CCAM downscaling of `NorESM2-MM` for 2005--2014 (historical) and
2090--2099 (`ssp245`, `ssp585`). A production study would use a multi-decade
reanalysis record and 20-year climate windows; the code is identical, and
the caveats at the end are largely about that difference in sample size.

## Baseline and change signal

The bias-corrected daily baseline and the point CMIP6 series, exactly as in
`vignette("scenario-workflow")`.

```{r baseline, message = FALSE, warning = FALSE}
obs <- prepare_obs_met(file.path(ex, "rotorua_buoy_met_aeme_hr.csv.gz"),
                       resample = "hour", tz = tz, station = "rotorua_buoy",
                       wind_height = 2, verbose = FALSE)

era5 <- read.csv(file.path(ex, "rotorua_era5_hourly_met.csv.gz"), check.names = FALSE)
era5$Date <- as.POSIXct(era5$Date, tz = tz, format = "%Y-%m-%d %H:%M:%S")
attr(era5, "tz") <- tz; attr(era5, "lat") <- lat; attr(era5, "lon") <- lon

fit_vars <- c("MET_tmpair", "MET_tmpdew", "MET_wndspd", "MET_radswd",
              "MET_humrel", "MET_prsttn")
bc <- fit_met_bias_correction(era5, obs, vars = fit_vars, method = "scale",
                              by = "doy-loess", verbose = FALSE)

## corrected hourly record (donor for disaggregation) and corrected daily baseline
era5_corr <- apply_met_bias_correction(era5, bc, expand = TRUE, lat = lat,
                                       lon = lon, elev = elev, tz = tz,
                                       verbose = FALSE)
baseline  <- bias_correct_daily_baseline(era5, bc, lat = lat, lon = lon,
                                         elev = elev, tz = tz, verbose = FALSE)
baseline  <- baseline[stats::complete.cases(
  baseline[c("MET_radswd", "MET_tmpair", "MET_pprain",
             "MET_wndspd", "MET_humrel")]), ]

cmip <- extract_cmip6_point(file.path(ex, "rotorua_cmip6"), lon = lon, lat = lat,
                            vars = c("MET_tmpair", "MET_pprain", "MET_wndspd",
                                     "MET_radswd", "MET_humrel"),
                            verbose = FALSE)
cmip_by   <- split(cmip, cmip$experiment)
REF_YEARS <- 2005:2014
FUT_YEARS <- 2090:2099
```

## What a mean delta-change does to extremes

Take the model's own wet-day rainfall (days with &ge; 1 mm) and compare the
future and historical distributions quantile by quantile.

```{r motivation}
wet <- function(x, mm = 1) x[is.finite(x) & x >= mm]

qq <- c(0.5, 0.75, 0.9, 0.95, 0.98, 0.99)
ratio <- quantile(wet(cmip_by$ssp585$MET_pprain), qq, type = 8) /
         quantile(wet(cmip_by$historical$MET_pprain), qq, type = 8)
round(ratio, 3)

## mean wet-day amount and wet-day frequency
c(mean_ratio = mean(wet(cmip_by$ssp585$MET_pprain)) /
              mean(wet(cmip_by$historical$MET_pprain)),
  wetday_ratio = sum(cmip_by$ssp585$MET_pprain >= 1, na.rm = TRUE) /
                sum(cmip_by$historical$MET_pprain >= 1, na.rm = TRUE))
```

The mean wet-day amount barely changes (ratio near 1) and there are
slightly *fewer* wet days, but the upper quantiles rise steeply --- the
model is projecting rain that falls less often and harder. A single mean
change factor applied to the observed series (`apply_delta()` in
`vignette("scenario-workflow")`) multiplies every quantile by that same
near-1 number and throws the tail signal away.

```{r motivation-fig, fig.height = 3, fig.cap = "**Figure 1.** Ratio of the `ssp585` 2090--2099 wet-day rainfall quantiles to the 2005--2014 model baseline, by non-exceedance probability. A mean delta-change (dashed) moves every quantile by the same factor; the model's own change (points) steepens with probability, so the rare heavy days grow while the median wet day shrinks."}
pr <- seq(0.5, 0.98, 0.02)
qr <- data.frame(
  p = pr,
  ratio = quantile(wet(cmip_by$ssp585$MET_pprain), pr, type = 8) /
          quantile(wet(cmip_by$historical$MET_pprain), pr, type = 8))
mr <- mean(wet(cmip_by$ssp585$MET_pprain)) / mean(wet(cmip_by$historical$MET_pprain))

ggplot(qr, aes(p, ratio)) +
  geom_hline(yintercept = mr, linetype = 2, colour = "grey45") +
  geom_hline(yintercept = 1, linewidth = 0.3, colour = "grey80") +
  geom_point(colour = "firebrick", size = 1.6) +
  labs(x = "wet-day non-exceedance probability",
       y = "ssp585 / historical")
```

## Option 1 --- quantile change factors

Instead of one factor per month, derive a change factor at several points
of the distribution and interpolate it onto each baseline day by that day's
rank. This is a quantile-perturbation / quantile-delta scheme: it keeps the
observed sequence, wet-day count and timing untouched, but lets the tail
stretch or contract relative to the centre.

```{r qcf}
PROB <- c(0.10, 0.25, 0.40, 0.55, 0.70, 0.80, 0.88, 0.94, 0.98)

## multiplicative change factors for wet-day rainfall, applied by rank
qcf_wetday <- function(base_rain, hist_rain, fut_rain,
                       probs = PROB, wet_mm = 1) {
  hh <- wet(hist_rain, wet_mm)
  ff <- wet(fut_rain,  wet_mm)
  cf <- quantile(ff, probs, names = FALSE, type = 8) /
        quantile(hh, probs, names = FALSE, type = 8)
  out <- base_rain
  i   <- which(is.finite(base_rain) & base_rain >= wet_mm)
  p   <- stats::ecdf(hh)(base_rain[i])                # rank of each wet day
  out[i] <- base_rain[i] * stats::approx(probs, cf, xout = p, rule = 2)$y
  structure(out, cf = data.frame(prob = probs, cf = cf))
}

proj_rain <- qcf_wetday(baseline$MET_pprain,
                        cmip_by$historical$MET_pprain,
                        cmip_by$ssp585$MET_pprain)
attr(proj_rain, "cf")
```

The warped baseline now tracks the modelled steepening far better than a
flat factor, though a ten-year sample leaves the very top quantile noisy
(the change factor is anchored no higher than the 98th percentile):

```{r qcf-check}
targ <- quantile(wet(cmip_by$ssp585$MET_pprain), qq, type = 8) /
        quantile(wet(cmip_by$historical$MET_pprain), qq, type = 8)
got  <- quantile(wet(proj_rain), qq, type = 8) /
        quantile(wet(baseline$MET_pprain), qq, type = 8)
round(rbind(model_target = targ, quantile_delta = got,
            mean_delta = rep(mr, length(qq))), 3)
```

```{r qcf-fig, fig.height = 3, fig.cap = "**Figure 2.** Wet-day rainfall exceedance curve for the corrected baseline and for the same days after quantile change factors. The median wet day is almost unchanged; the upper tail is lifted, following the model signal in Figure 1. Timing and the number of wet days are exactly those of the baseline."}
exc <- function(x, lab) {
  x <- sort(wet(x), decreasing = TRUE)
  data.frame(value = x, p_exceed = seq_along(x) / (length(x) + 1), series = lab)
}
ec <- rbind(exc(baseline$MET_pprain, "baseline"),
            exc(proj_rain,           "+ quantile delta (ssp585)"))

ggplot(ec, aes(value, p_exceed, colour = series)) +
  geom_line() +
  scale_y_log10() +
  scale_colour_manual(values = c("baseline" = "grey55",
                                 "+ quantile delta (ssp585)" = "firebrick")) +
  labs(x = "daily rainfall (mm)", y = "exceedance probability (log)")
```

The same idea covers hot extremes: split the monthly air-temperature change
into a lower-quantile and an upper-quantile shift so heatwave days warm
more (or less) than an average day, rather than moving the whole month
rigidly.

## Option 2 --- drive from bias-corrected model output

Quantile change factors still ride on the observed calendar, so they cannot
add wet days, lengthen a drought, or move a storm to another month. To get
those you have to take the model's *own* daily sequence. The bundled
`tas` and `pr` files are distributed already bias-corrected to a 5 km
observational grid (`_bc` in the file names); `hurs`, `rsds` and `sfcWind`
are raw. A workable hybrid is: use the corrected model series directly for
temperature and rainfall, and delta-change the rest onto the baseline.

```{r hybrid}
seas_delta <- function(v, kind) {
  h <- cmip_by$historical; f <- cmip_by$ssp585
  mh <- as.integer(format(h$Date, "%m")); hy <- as.integer(format(h$Date, "%Y"))
  mf <- as.integer(format(f$Date, "%m")); fy <- as.integer(format(f$Date, "%Y"))
  a <- tapply(h[[v]][hy %in% REF_YEARS], mh[hy %in% REF_YEARS], mean, na.rm = TRUE)
  b <- tapply(f[[v]][fy %in% FUT_YEARS], mf[fy %in% FUT_YEARS], mean, na.rm = TRUE)
  d <- if (kind == "ratio") b / a else b - a
  stats::setNames(as.numeric(d[as.character(1:12)]), 1:12)
}

mo     <- as.integer(format(baseline$Date, "%m"))
hybrid <- baseline
hybrid$MET_tmpair <- baseline$MET_tmpair + seas_delta("MET_tmpair", "add")[mo]
hybrid$MET_radswd <- baseline$MET_radswd + seas_delta("MET_radswd", "add")[mo]
hybrid$MET_wndspd <- baseline$MET_wndspd * seas_delta("MET_wndspd", "ratio")[mo]
hybrid$MET_humrel <- baseline$MET_humrel * seas_delta("MET_humrel", "ratio")[mo]
hybrid$MET_pprain <- as.numeric(proj_rain)
hybrid <- expand_met(hybrid[, c("Date", "MET_radswd", "MET_tmpair", "MET_pprain",
                                "MET_humrel", "MET_wndspd", "MET_prsttn")],
                     lat = lat, lon = lon, elev = elev, tz = tz)
```

Whether or not you route temperature and rainfall through the hybrid, the
change in event *frequency and clustering* only exists in the model
series. Dry-spell lengths, model historical vs `ssp585`:

```{r spells}
spell_len <- function(is_dry) { r <- rle(is_dry); r$lengths[r$values] }
dh <- spell_len(cmip_by$historical$MET_pprain < 1)
df <- spell_len(cmip_by$ssp585$MET_pprain     < 1)
rbind(historical = c(mean = mean(dh), p90 = quantile(dh, 0.9, names = FALSE), max = max(dh)),
      ssp585     = c(mean = mean(df), p90 = quantile(df, 0.9, names = FALSE), max = max(df)))
```

```{r spells-fig, fig.height = 3, fig.cap = "**Figure 3.** Distribution of consecutive dry-day (< 1 mm) run lengths in the model historical and `ssp585` windows. The bulk is similar but the `ssp585` upper tail reaches longer droughts. A delta-change or quantile-delta projection keeps the observed run-length distribution unchanged and cannot show this."}
sp <- rbind(data.frame(len = dh, exp = "historical"),
            data.frame(len = df, exp = "ssp585"))
ggplot(sp, aes(len, colour = exp)) +
  stat_ecdf() +
  coord_cartesian(xlim = c(0, max(df))) +
  scale_colour_manual(values = c(historical = "grey45", ssp585 = "firebrick")) +
  labs(x = "dry-spell length (days)", y = "ECDF")
```

The price is that you inherit the model's day-to-day behaviour and its
sub-daily silence: the series is daily, so hourly structure still has to
come from a donor (next section), and any error in the model's wet-day
sequencing is now in your forcing.

## Option 3 --- storyline perturbation of an observed event

When the study is about one kind of event --- the flood-driving storm, the
stratification-breaking gale --- perturb a real observed event with the
projected thermodynamics and drive the model with that. Here: take the
wettest seven-day window of the corrected record, warm it by the seasonal
mean change, scale its rainfall by Clausius--Clapeyron (about 7% per degree
of warming), scale wind by its own change factor, then disaggregate the
perturbed daily event to hourly.

```{r storyline}
dd    <- as.Date(era5_corr$Date, tz = tz)
wksum <- tapply(era5_corr$MET_pprain, dd, sum, na.rm = TRUE)
udays <- as.Date(names(wksum))
roll7 <- as.numeric(stats::filter(as.numeric(wksum), rep(1, 7), sides = 1))
d0    <- udays[which.max(roll7)] - 6

ev_hr  <- era5_corr[dd >= d0 & dd < d0 + 7, ]
ev_day <- met_to_daily(ev_hr, tz = tz, min_frac = 1)
m      <- as.integer(format(d0, "%m"))

dT  <- seas_delta("MET_tmpair", "add")[m]        # seasonal mean warming
fW  <- seas_delta("MET_wndspd", "ratio")[m]
a   <- 0.07                                       # CC scaling per degC

story <- ev_day
story$MET_tmpair <- ev_day$MET_tmpair + dT
story$MET_pprain <- ev_day$MET_pprain * (1 + a)^dT
story$MET_wndspd <- ev_day$MET_wndspd * fW
story <- expand_met(story[, c("Date", "MET_radswd", "MET_tmpair", "MET_pprain",
                              "MET_humrel", "MET_wndspd", "MET_prsttn")],
                    lat = lat, lon = lon, elev = elev, tz = tz)

round(c(warming_degC = unname(dT), rain_factor = unname((1 + a)^dT),
        wind_factor  = unname(fW)), 3)

story_hr <- disaggregate_met_to_hourly(story, donor = era5_corr,
                                       method = "fragments", swr = "clearsky",
                                       lat = lat, lon = lon, elev = elev, tz = tz,
                                       seed = 1, expand = TRUE, verbose = FALSE)
```

```{r storyline-fig, fig.height = 4, fig.cap = "**Figure 4.** The wettest observed week in the corrected record (grey) and its storyline `ssp585` counterpart (red): air temperature shifted by the seasonal mean warming, hourly rainfall scaled by Clausius--Clapeyron, wind by its seasonal factor. The event keeps its observed timing and within-day structure; its intensity is rescaled."}
vs <- c(MET_tmpair = "air temperature (°C)",
        MET_pprain = "rainfall (mm/hour)",
        MET_wndspd = "wind speed (m/s)")
mk <- function(d, lab) data.frame(
  Date = rep(d$Date, length(vs)),
  value = unlist(d[names(vs)], use.names = FALSE),
  variable = factor(rep(unname(vs), each = nrow(d)), unname(vs)),
  series = lab)
sf <- rbind(mk(ev_hr, "observed"), mk(story_hr, "storyline ssp585"))

ggplot(sf, aes(Date, value, colour = series)) +
  geom_line(linewidth = 0.4) +
  facet_wrap(~ variable, ncol = 1, scales = "free_y", strip.position = "left") +
  scale_colour_manual(values = c("observed" = "grey55",
                                 "storyline ssp585" = "firebrick")) +
  labs(x = NULL, y = NULL)
```

Storylines make no frequency claim --- they answer "how would *this* event
look in a warmer climate", not "how often". Run a handful spanning the
observed event set rather than one.

## Disaggregation settings that matter for extremes

Any daily projection still has to be disaggregated for a sub-daily lake
model. For extremes the defaults of `disaggregate_met_to_hourly()` need a
second look.

* **`method = "fragments"`** (the default) borrows a whole observed day's
  shape, so a wet target day gets a genuinely intermittent wet-hour
  pattern. `method = "diurnal"` imposes a smooth mean cycle and will
  understate peak hourly rain and gusts --- avoid it when the sub-daily
  peak is the point.
* **`swr = "clearsky"`** rebuilds shortwave from solar geometry, so a
  bright clearing after a front is not damped by an analogue day. Keep it.
* **`sample_top_k` and `seed`** control how much the borrowed shape varies.
  With `sample_top_k > 1` the donor is drawn from the *k* closest analogue
  days, so a seed ensemble maps the sub-daily uncertainty that a single run
  hides:

```{r seed-ens}
wm    <- names(which.max(tapply(hybrid$MET_pprain, format(hybrid$Date, "%Y-%m"), sum)))
slice <- hybrid[format(hybrid$Date, "%Y-%m") == wm, ]

peak <- t(sapply(1:12, function(s) {
  h <- disaggregate_met_to_hourly(slice, donor = era5_corr, method = "fragments",
         swr = "clearsky", lat = lat, lon = lon, elev = elev, tz = tz,
         seed = s, sample_top_k = 5, expand = FALSE, verbose = FALSE)
  c(peak_rain = max(h$MET_pprain), peak_wind = max(h$MET_wndspd))
}))
round(rbind(min = apply(peak, 2, min), max = apply(peak, 2, max)), 2)
sum(slice$MET_pprain)   # daily totals are identical across every seed
```

The peak hourly wind of the month swings by several m s^-1^ between seeds
while every daily mean is held fixed. The peak hourly *rain* here barely
moves, for a revealing reason: the donor record is only three years long,
so the very wettest projected days have almost no close analogue and are
disaggregated from essentially one donor day. A longer donor record is
what widens that pool --- see the caveats.

* **Conservation.** Disaggregation conserves daily means and daily totals,
  so it never invents a daily extreme that was not in the projection ---
  and never sharpens a sub-hourly burst beyond what the donor day had.

```{r conservation}
hh   <- disaggregate_met_to_hourly(slice, donor = era5_corr, method = "fragments",
          swr = "clearsky", lat = lat, lon = lon, elev = elev, tz = tz,
          seed = 1, expand = TRUE, verbose = FALSE)
back <- met_to_daily(hh, tz = tz, min_frac = 1)
i    <- match(as.Date(back$Date), as.Date(slice$Date))
sapply(c("MET_tmpair", "MET_humrel", "MET_pprain"),
       function(v) max(abs(back[[v]] - slice[[v]][i])))
```

Wind speed is the exception: it is conserved to the daily *vector* mean,
not the scalar mean speed that `met_to_daily()` reports, so the two differ
by the within-day directional variability (see
`?disaggregate_met_to_hourly`). That is a property of the wind convention,
not a disaggregation error.

## Diagnosing extremes in the forcing

Before a model run, confirm the forcing actually carries the extremes you
expect. Cheap checks:

```{r diag}
annual_max <- function(d, v) {
  yr <- as.integer(format(d$Date, "%Y"))
  tapply(d[[v]], yr, max, na.rm = TRUE)
}
rbind(historical = summary(annual_max(cmip_by$historical, "MET_pprain")),
      ssp585     = summary(annual_max(cmip_by$ssp585,     "MET_pprain")))

## wet-hour fraction of the disaggregated projection vs the donor
c(donor   = mean(era5_corr$MET_pprain > 0.1),
  project = mean(story_hr$MET_pprain > 0.1))
```

Useful diagnostics, roughly in order of effort: annual-block maxima of
daily rain and daily-mean wind; empirical return periods (rank the values,
plot value against `(n + 1) / rank / years`); wet-day and wet-hour
fractions; dry- and calm-spell length distributions (`rle()` as above);
and, if the model output allows it, the seasonal timing of the annual
maximum. Compare every one of these between baseline and projection, and
between disaggregated hourly and its daily input.

## Handing extreme forcing to a lake model

`met_to_cf()` writes the projection to a CF-compliant table; with `AEME`
installed it goes straight onto an `aeme` object.

```{r aeme, eval = FALSE}
inp <- AEME::input(aeme)
inp$meteo <- story_hr          # or the disaggregated hybrid projection
AEME::input(aeme) <- inp
```

Points specific to an extremes study:

* **Match the driver to the mechanism.** Mixing and surface heat/moisture
  fluxes scale with wind speed squared or cubed, so they are set by the
  sub-daily gust, not the daily mean --- the seed ensemble above is part of
  the forcing uncertainty. Heavy rain usually enters a lake model through a
  separate catchment-hydrology step; the direct surface term is the
  freshwater lens and the latent-heat change. Stratification and its
  breakdown follow shortwave and the cold-outbreak minima.
* **Timestep.** Drive at the hourly resolution the disaggregator produces;
  let the model sub-step. Aggregating back to daily before the run discards
  the reason for doing any of this.
* **Spin-up.** Precede each window with one to two years of forcing so deep
  water and sediment temperature are not carrying the initial condition
  into the event of interest.
* **Ensemble, not a run.** The irreducible spread is
  disaggregation seed x scenario x (ideally) GCM. A single projected
  hourly series is one draw; return periods and event magnitudes need the
  spread.
* **Conserve and check.** Re-aggregate the hourly forcing with
  `met_to_daily()` and confirm it reproduces the daily projection before
  trusting any model output built on it.

## Caveats

* **Stationarity, again.** Every method here assumes a relationship fitted
  on the past --- a quantile map, a change factor, a CC rate, a donor
  library --- holds in the future. The assumption is strongest for the mean
  and weakens quickly into the tail.
* **Quantile change factors and delta-change keep the observed calendar.**
  They can restretch the distribution but cannot add an event, lengthen a
  spell, or move a storm's season. Only Option 2 can, and only to the
  extent the model gets sequencing right.
* **Short records starve the tail.** Ten-year windows put only a handful of
  values above the 98th percentile, so the change factors and return
  periods in this vignette are illustrative, not robust --- use long
  reanalysis records and 20-year (or pooled multi-model) windows. A short
  hourly donor likewise leaves the rarest days with one analogue; lengthen
  it to sample sub-daily structure.
* **Disaggregation borrows, it does not sharpen.** The hourly series can be
  no peakier than the observed donor days it is built from; genuine change
  in sub-daily intensity (convective bursts under warming) is outside what
  a method-of-fragments disaggregator can produce.
* **One model is not a projection.** Treat the `ssp245`--`ssp585` spread as
  a lower bound on uncertainty and add more GCMs before quoting a number.

See `?scenario_workflow` and `vignette("scenario-workflow")` for the
mean-change pipeline these methods extend.
