---
title: "Deriving a complete met set"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Deriving a complete met set}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 90)
```

A lake model wants a dozen or more meteorological variables, but a station
or a reanalysis extract usually gives you five or six. `expand_met()`
derives the rest from a small required core, filling in **only what is
missing**. This vignette shows `expand_met()` and then each of the
underlying conversions on its own, because their assumptions are what
determine whether the derived values are any good.

```{r setup}
library(metscale)
ex   <- system.file("extdata", package = "metscale")
lat  <- -38.0790
lon  <- 176.2717
elev <- 279          # Lake Rotorua surface, m
```

## `expand_met()`

The required core is: `MET_radswd` (W m-2), `MET_tmpair` (degC),
`MET_pprain` (mm per timestep), **one** of `MET_humrel` (%) or
`MET_tmpdew` (degC), and wind as either `MET_wndspd` or both
`MET_wnduvu` / `MET_wnduvv`. Take a minimal hourly slice:

```{r minimal}
era5 <- read.csv(file.path(ex, "rotorua_era5_hourly_met.csv.gz"), check.names = FALSE)
era5$Date <- as.POSIXct(era5$Date, tz = "Etc/GMT-12")

minimal <- era5[1:72, c("Date", "MET_radswd", "MET_tmpair", "MET_pprain",
                        "MET_humrel", "MET_wndspd")]
names(minimal)
```

```{r expand}
full <- expand_met(minimal, lat = lat, lon = lon, elev = elev,
                   tz = "Etc/GMT-12")
names(full)
setdiff(names(full), names(minimal))   # what expand_met() added
```

Nothing is resampled and supplied columns are never overwritten --- if you
pass `MET_cldcvr`, that is what comes back. `tz` is used only for the solar
geometry behind cloud cover, so pass the zone your `Date` column is
actually in.

## Humidity

Dew point, relative humidity and vapour pressure are three views of the
same quantity; give any one (plus air temperature) and the others follow.
`metscale` uses Magnus--Tetens with the Alduchov & Eskridge coefficients.

```{r humidity}
airt <- c(5, 15, 25)
rh   <- c(90, 60, 40)

td <- rh_to_dewpoint(airt, rh)
td
dewpoint_to_rh(airt, td)          # inverse
vapour_pressure(airt, rh)         # actual vapour pressure, hPa
```

## Pressure

`station_pressure()` is the hypsometric equation with virtual temperature
(it accounts for the humidity dependence of air density), reducing a
sea-level pressure to the lake surface. `mslp_from_station()` /
`station_from_mslp()` convert the other way.

```{r pressure}
ps <- station_pressure(airt = 12, relh = 80, elev = elev)   # Pa
ps
mslp_from_station(ps, elev = elev, airt = 12)               # back to sea level
```

This is why a raw ERA5-Land `MET_prsttn` can look ~3 kPa low at a lake: the
reanalysis reports pressure at its grid-cell orography height, which is
often a few hundred metres above the actual water surface. Supplying the
true `elev` to `expand_met()` (or bias-correcting the pressure) fixes it.

## Wind vector

`ds2uv()` and `uv2ds()` are mutual inverses using the meteorological
convention --- direction is where the wind comes **from**.

```{r wind}
uv <- ds2uv(d = 225, s = 5)      # SW wind at 5 m/s
uv
uv2ds(uv[, "u"], uv[, "v"])      # -> dir 225, speed 5
```

(The AEME port fixed a u/v swap here; `metscale`'s components are
consistent with its own speed/direction.) To move a measured wind between
sensor heights, use `wind_at_height()` / `met_wind_at_height()` --- a 2 m
buoy anemometer reads roughly 15--20% low against the 10 m convention.

## Cloud cover from measured shortwave

`calc_cc()` inverts the ratio of measured to clear-sky potential shortwave:
a bright day implies little cloud, a dim one implies a lot. It needs the
timestamps, air temperature, dew point (or RH), position and `tz`.

```{r cloud}
cc <- calc_cc(minimal$Date, airt = minimal$MET_tmpair, swr = minimal$MET_radswd,
              relh = minimal$MET_humrel, lat = lat, lon = lon, elev = elev,
              tz = "Etc/GMT-12")
summary(cc)                       # 0-1 fraction
```

Because it is a daytime ratio, night-time values are interpolated and a
day with no usable daylight (or with `MET_radswd` already an all-day mean)
gives a weak estimate.

## Longwave

`calc_in_lwr()` builds downwelling longwave from a clear-sky emissivity
(Idso & Jackson) with a cloud correction:

```{r longwave}
calc_in_lwr(cc = c(0, 0.5, 1), airt = c(10, 10, 10), relh = c(80, 80, 80))
```

## Solar geometry and shortwave

```{r solar}
noon <- as.POSIXct("2024-01-15 13:00", tz = "Etc/GMT-12")
solar_zenith_angle(noon, lat = lat, lon = lon)            # degrees from vertical
clear_sky_swr(noon, lat = lat, lon = lon)                 # W m-2, cloud-free
```

`estimate_hourly_swr()` goes from a **daily-mean** `MET_radswd` to an
hourly curve with real night-time zeros, a peak at local solar noon, and
each day's mean preserved:

```{r hourly-swr}
day <- data.frame(Date = as.Date("2024-01-15") + 0:2,
                  MET_radswd = c(280, 120, 310))
hr  <- estimate_hourly_swr(day, lat = lat, lon = lon, tz = "Etc/GMT-12")

## recover the daily means -- aggregating in the SAME zone the series is
## labelled in; `as.Date()` defaults to UTC and would split days at the
## wrong instant
tapply(hr$MET_radswd, as.Date(hr$Date, tz = "Etc/GMT-12"), mean)   # == the inputs
hr[which.max(hr$MET_radswd), ]                                     # peak near local noon
```

## Three assumptions worth remembering

These are the behaviours `metscale` deliberately changed from the AEME
originals, because getting them wrong is silent:

* **Time zone.** `expand_met()`, `calc_cc()` and `estimate_hourly_swr()`
  phase the solar geometry to `tz`. It defaults to the `tz` attribute of
  the input, else `"UTC"`; pass the zone your timestamps are in (e.g.
  `"Etc/GMT-12"` for fixed NZST) to put the shortwave peak at local noon.
* **Pressure units.** Derived `MET_prsttn` / `MET_prmslp` are in **Pa**.
* **Wind components.** `ds2uv()` / `uv2ds()` round-trip; derived u/v are
  consistent with the speed/direction they came from.

## See also

* `vignette("met-data-frames")` --- the names, units and time-zone
  convention these functions assume.
* `vignette("scenario-workflow")` --- `expand_met()` in a full projection
  pipeline, where it regenerates dependents after each adjustment.
