---
title: "Download ERA5 data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Download ERA5 data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

```{r setup}
library(metscale)
```

`metscale`'s bias-correction and disaggregation tools operate on a meteorology
data frame. This vignette covers the three ways to *acquire* ERA5 forcing, from
lightest to heaviest. Whichever you use, the output feeds into
`standardise_met()`, `fit_met_bias_correction()` and
`disaggregate_met_to_hourly()`.

All three functions need suggested packages that are **not** installed with
`metscale` by default:

```{r install-deps, eval=FALSE}
install.packages(c("httr2", "jsonlite", "terra", "ecmwfr", "stars", "sf"))
```

## 1. Point time series from ISIMIP3a — `download_era5_isimip_point()`

The quickest route. `download_era5_isimip_point()` pulls daily ERA5
(20CRv3-ERA5, ISIMIP3a `obsclim`) from the
[ISIMIP repository API](https://files.isimip.org/api/v2) for any point on the
globe. No account or key is required. Coverage currently runs to 2021.

```{r isimip}
lon <- 98.67591   # Lake Toba, Indonesia
lat <- 2.637047
years <- 2015:2021
vars <- c("MET_tmpair", "MET_pprain")

met <- download_era5_isimip_point(lon, lat, years, vars)
summary(met)
```

Variable names may be given either as AEME `MET_*` names or as the ERA5/CMIP
short names (`tas`, `pr`, `sfcwind`, `rsds`, `ps`, `rlds`, `hurs`). The result
has a `Date` column plus one column per variable, already in AEME names and
units (degC, mm/day, ...).

## 2. Gridded GRIB from the Copernicus Data Store — `download_era5_cds()`

For the full hourly ERA5 / ERA5-Land archive you need a free
[ECMWF account](https://www.ecmwf.int/user/login) or a free
[Copernicus Data Store (CDS)](https://cds.climate.copernicus.eu/user/register)
account, linked to your session with `ecmwfr::wf_set_key()`. See the
[ecmwfr documentation](https://bluegreen-labs.github.io/ecmwfr/#use).

```{r cds-key, eval=FALSE}
ecmwfr::wf_set_key(key = Sys.getenv("ECMWF_KEY"),
                   user = Sys.getenv("ECMWF_USER"))
```

`download_era5_cds()` submits one CDS request per variable / year / month
(batched, up to 20 at a time) and writes a GRIB file per request. Give it a
point plus `buffer`, or an `sf` polygon as `shape`.

File downloads can be slow, so this vignette uses a pre-downloaded GRIB for
demonstration. The `extdata` directory contains a single GRIB for Lake Toba,
Indonesia (2m temperature, January 2024).


```{r echo=FALSE}
files <- system.file("extdata/era5/reanalysis-era5-land_2m_temperature_hourly_2024_1_toba.grib",
                     package = "metscale")
year <- 2024
month <- 1
```

```{r cds-download, message=FALSE, eval=FALSE}
year <- 2024
month <- 1
files <- download_era5_cds(
  lat = lat, lon = lon, year = year, month = month,
  variable = "2m_temperature", path = "data/test", site = "toba",
  user = Sys.getenv("ECMWF_USER")
)
```
```{r}
files
```

Plot the downloaded GRIB with `plot_era5_grib()`:

Plot the point location on the downloaded grid with `plot_extract_grid()`:

```{r}
#| fig.width: 6
#| fig.height: 6
#| fig.caption: "ERA5 grid and extraction point"
path <- dirname(files)
plot_extract_grid(path = path, lon = lon, lat = lat, method = "bilinear")
```


Extract a point (or polygon-mean) time series from the downloaded GRIB with
`read_era5_grib_point()`:

```{r grib-point}
df <- read_era5_grib_point(file = files, lat = lat, lon = lon)
head(df)
```

For a model-ready hourly frame (de-accumulated fluxes, standard units, a
time-zone shift) point `extract_era5_hourly_met()` at the download
directory instead. It picks the reader from each file's extension, so the
GRIB written above is read directly; the default `pattern` (`"{variable}"`)
already matches `download_era5_cds()` names, so nothing else is needed:

```{r grib-hourly}
met <- extract_era5_hourly_met(
  path = path, lon = lon, lat = lat, years = year, variables = "2m_temperature"
  )
```

## 3. Aggregate downloaded ERA5 to daily — `convert_era5_netcdf()`

For a **daily** frame, `convert_era5_netcdf()` wraps the step above: it runs
`extract_era5_hourly_met()` over the files in `path` (netCDF or GRIB — the
same file discovery, spatial sampling, de-accumulation, unit conversion and
time-zone shift) and then aggregates to daily with `met_to_daily()` (rain and
snow summed, everything else averaged). Extra `...` arguments are forwarded to
`extract_era5_hourly_met()`.

```{r convert}
met <- convert_era5_netcdf(
  path = path, lon = lon, lat = lat, years = year,
  variables = "2m_temperature", site = "toba", format = "AEME"
)
```

Unless `minmax = FALSE`, daily minimum and maximum air and dewpoint
temperature are added (`MET_airmin` / `MET_airmax` / `MET_dewmin` /
`MET_dewmax`).

`convert_era5_netcdf(format = "raw")` returns the ERA5 short names
(`t2m`, `d2m`, `ssrd`, `strd`, ...) with a `datetime` column, which
`standardise_met()` recognises directly:

```{r standardise}
met_raw <- convert_era5_netcdf(path = path, lon = lon, lat = lat, years = year,
                               variables = "2m_temperature", format = "raw")
std <- standardise_met(met_raw)
```

## Next steps

- `standardise_met()` — column names, units, time zone, resampling.
- `fit_met_bias_correction()` / `apply_met_bias_correction()` — correct ERA5
  against local observations.
- `disaggregate_met_to_hourly()` — daily back to sub-daily.
- `expand_met()` — derive the remaining variables a lake model needs.
