---
title: "Download ERA5 data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Download ERA5 data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(climprep)
```

`climprep`'s bias-correction and disaggregation tools operate on a meteorology
data frame. This vignette covers the three ways to *acquire* ERA5 forcing, from
lightest to heaviest. Whichever you use, the output feeds into
`standardise_met()`, `fit_met_bias_correction()` and
`disaggregate_met_to_hourly()`.

All three functions need suggested packages that are **not** installed with
`climprep` by default:

```{r install-deps}
install.packages(c("httr2", "jsonlite", "terra", "ecmwfr", "stars", "sf"))
```

## 1. Point time series from ISIMIP3a — `download_era5_isimip_point()`

The quickest route. `download_era5_isimip_point()` pulls daily ERA5
(20CRv3-ERA5, ISIMIP3a `obsclim`) from the
[ISIMIP repository API](https://files.isimip.org/api/v2) for any point on the
globe. No account or key is required. Coverage currently runs to 2021.

```{r isimip}
lon <- 98.67591   # Lake Toba, Indonesia
lat <- 2.637047
years <- 2015:2021
vars <- c("MET_tmpair", "MET_pprain")

met <- download_era5_isimip_point(lon, lat, years, vars)
summary(met)
```

Variable names may be given either as AEME `MET_*` names or as the ERA5/CMIP
short names (`tas`, `pr`, `sfcwind`, `rsds`, `ps`, `rlds`, `hurs`). The result
has a `Date` column plus one column per variable, already in AEME names and
units (degC, mm/day, ...).

## 2. Gridded GRIB from the Copernicus Data Store — `download_era5_cds()`

For the full hourly ERA5 / ERA5-Land archive you need a free
[Copernicus Data Store (CDS)](https://cds.climate.copernicus.eu/user/register)
account, linked to your session with `ecmwfr::wf_set_key()`. See the
[ecmwfr documentation](https://bluegreen-labs.github.io/ecmwfr/#use).

```{r cds-key}
ecmwfr::wf_set_key(key = Sys.getenv("CDS_KEY"), user = Sys.getenv("CDS_USER"))
```

`download_era5_cds()` submits one CDS request per variable / year / month
(batched, up to 20 at a time) and writes a GRIB file per request. Give it a
point plus `buffer`, or an `sf` polygon as `shape`.

```{r cds-download}
year <- 2024
month <- 1
files <- download_era5_cds(
  lat = lat, lon = lon, year = year, month = month,
  variable = "2m_temperature", path = "data/test", site = "toba",
  user = Sys.getenv("CDS_USER")
)
files
```

Extract a point (or polygon-mean) time series from the downloaded GRIB with
`read_era5_grib_point()`:

```{r grib-point}
df <- read_era5_grib_point(file = files, lat = lat, lon = lon)
head(df)
```

## 3. Aggregate downloaded netCDF to daily — `convert_era5_netcdf()`

If you already have hourly ERA5 netCDF on disk (named
`era5_<variable>_hourly_<year>_<site>.nc`), `convert_era5_netcdf()` aggregates
each variable to daily with the appropriate function (mean, or max for
precipitation / radiation) and returns AEME or LakeEnsemblR columns.

```{r convert}
met <- convert_era5_netcdf(
  lat = lat, lon = lon, year = 2022:2023,
  path = "era5_netcdf", site = "toba", format = "AEME"
)
```

For **hourly** extraction from ERA5-Land netCDF — with de-accumulation of the
flux variables, unit conversion and a time-zone shift — use
`extract_era5_hourly_met()` / `extract_era5_lake_met()` instead.

`convert_era5_netcdf(format = "raw")` returns the ERA5 nc short names
(`t2m`, `d2m`, `ssrd`, `strd`, ...), which `standardise_met()` recognises
directly:

```{r standardise}
met_raw <- convert_era5_netcdf(lat = lat, lon = lon, year = 2022,
                               path = "era5_netcdf", format = "raw")
std <- standardise_met(met_raw)
```

## Next steps

- `standardise_met()` — column names, units, time zone, resampling.
- `fit_met_bias_correction()` / `apply_met_bias_correction()` — correct ERA5
  against local observations.
- `disaggregate_met_to_hourly()` — daily back to sub-daily.
- `expand_met()` — derive the remaining variables a lake model needs.
