---
title: "Extracting ERA5-Land over a lake"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extracting ERA5-Land over a lake}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
options(width = 90)
```

Once you have hourly ERA5-Land netCDF on disk (see
`vignette("download-era5")`), `extract_era5_hourly_met()` pulls a
`metscale`-shaped hourly series out of it for a point or a polygon.
`extract_era5_lake_met()` is a thin wrapper that looks a lake up in a
polygon layer first. The code chunks here are **not executed** (they need
the netCDF archive); the one exception is reading the bundled lake polygon.

```{r setup, eval = TRUE}
library(metscale)
ex <- system.file("extdata", package = "metscale")
```

## Point extraction

```{r point}
met <- extract_era5_hourly_met(
  path      = "era5_netcdf",          # directory of monthly hourly .nc files
  lon       = 176.2717,
  lat       = -38.0790,
  years     = 2023:2025,
  method    = "bilinear",             # 4 nodes around the point
  tz        = "Etc/GMT-12",
  format    = "AEME")                 # MET_* names and units
```

Files are located with `file_template`, an `sprintf()` pattern taking the
variable and year; the default matches the naming that
`download_era5_cds()` writes. `variables` defaults to the nine ERA5-Land
fields `metscale` needs; `precip_units` and `pressure_units` control the
output units (`"mm/hr"` and `"Pa"` by default). `fill_gaps = TRUE`
reindexes onto a complete regular hourly sequence.

## Polygon (lake-average) extraction

A lake wider than an ERA5-Land grid cell (~0.1&deg;, roughly 9 km at this
latitude) covers several cells, and a point sample then depends on exactly
where the point falls. Averaging over the cells that overlap the lake
polygon is more stable. Pass the geometry as `geom`:

```{r polygon}
lake_poly <- readRDS(file.path(ex, "rotorua_lake_shape.rds"))   # an sf polygon

met_area <- extract_era5_hourly_met(
  path   = "era5_netcdf",
  geom   = lake_poly,
  years  = 2023:2025,
  method = "area",                    # area-weighted mean of overlapping cells
  tz     = "Etc/GMT-12")
```

`method` chooses the spatial sampling:

| `method` | what it does | use when |
|---|---|---|
| `"nearest"` | single closest cell centre | lake well inside one cell |
| `"bilinear"` | 4 cell centres around the centroid | point-like water body (default) |
| `"area_mean"` | unweighted mean of cells overlapping the polygon | lake spans a few cells |
| `"area"` | **area-weighted** mean of overlapping cells | lake spans several cells unevenly |

The bundled Lake Rotorua polygon spans more than one cell, so `"area"` is
the right choice:

```{r inspect-shape, eval = requireNamespace("sf", quietly = TRUE)}
lake_poly <- readRDS(file.path(ex, "rotorua_lake_shape.rds"))
sf::st_crs(lake_poly)$epsg          # 2193 (NZTM); any CRS is accepted
round(as.numeric(sum(sf::st_area(lake_poly))) / 1e6, 1)   # lake area, km^2
sf::st_bbox(sf::st_transform(lake_poly, 4326))            # ~0.11 deg E-W
```

An ~0.11&deg; east--west extent against a ~0.1&deg; grid means the lake
touches two to three cells in each direction --- exactly the case
area-weighting is for.

## Looking a lake up by name

If you have a lake polygon *layer* (many lakes), `extract_era5_lake_met()`
matches one by id or name and forwards its geometry. It matches against
`id_final`, `id_LID`, `name_final`, `name_fenz`, `name_english` and
`name_maori` (case-insensitive).

```{r lake-lookup}
lakes <- readRDS("gis/lernzmp_lakes_master.rds")   # sf layer, or list of layers

met <- extract_era5_lake_met(
  lake   = "Rotorua",
  path   = "era5_netcdf",
  lakes  = lakes,
  method = "area",
  years  = 2023:2025,
  outfile = "rotorua_era5_hourly_met.csv")

## batch over several lakes
ids  <- c("LID 11133", "LID 25994", "LID 54732")
mets <- lapply(ids, extract_era5_lake_met, path = "era5_netcdf",
               lakes = lakes, method = "area", years = 2024)
```

The bundled `rotorua_era5_hourly_met.csv.gz` used throughout the other
vignettes is the output of exactly this call for Lake Rotorua.

## See also

* `vignette("download-era5")` --- getting the netCDF in the first place.
* `vignette("met-data-frames")` --- the `MET_*` names, units and time-zone
  convention the extractor output follows.
* `vignette("bias-correction-methods")` --- correcting the extracted series
  against local observations.
