---
title: "Extracting ERA5-Land over a lake"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extracting ERA5-Land over a lake}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
options(width = 90)
```

Once you have hourly ERA5-Land netCDF or GRIB on disk (see
`vignette("download-era5")`), `extract_era5_hourly_met()` pulls a
`metscale`-shaped hourly series out of it for a point or a polygon.
`extract_era5_lake_met()` is a thin wrapper that takes a lake polygon and
defaults to the area-weighted method. The extraction chunks here are **not
executed** (they need the reanalysis archive); the exceptions are reading
the bundled lake polygon and the grid visualisations near the end, which
run against a small synthesised grid.

```{r setup, eval = TRUE}
library(metscale)
ex <- system.file("extdata", package = "metscale")
```

## Point extraction

```{r point}
met <- extract_era5_hourly_met(
  path      = "era5_netcdf",          # directory of monthly hourly .nc files
  lon       = 176.2717,
  lat       = -38.0790,
  years     = 2023:2025,
  method    = "bilinear",             # 4 nodes around the point
  tz        = "Etc/GMT-12",
  format    = "AEME")                 # MET_* names and units
```

Files are located by matching `pattern` against the names in `path`. The
default, `"{variable}"`, finds any `.nc` / `.grib` file whose name holds
the ERA5 variable name and a 4-digit year -- enough for
`download_era5_cds()` output and most ad-hoc layouts. Give it explicitly
to disambiguate or to filter by month, e.g.
`pattern = "{year}_{month}_{variable}"` or
`pattern = "*_{variable}_hourly_{year}_{month}_*"`; `{year}` / `{month}`
are optional and `*` is a wildcard. `variables` defaults to the nine
ERA5-Land fields `metscale` needs; `precip_units` and `pressure_units`
control the output units (`"mm/hr"` and `"Pa"` by default).
`fill_gaps = TRUE` reindexes onto a complete regular hourly sequence.

## Polygon (lake-average) extraction

A lake wider than an ERA5-Land grid cell (~0.1&deg;, roughly 9 km at this
latitude) covers several cells, and a point sample then depends on exactly
where the point falls. Averaging over the cells that overlap the lake
polygon is more stable. Pass the geometry as `geom`:

```{r polygon}
lake_poly <- readRDS(file.path(ex, "rotorua_lake_shape.rds"))   # an sf polygon

met_area <- extract_era5_hourly_met(
  path   = "era5_netcdf",
  geom   = lake_poly,
  years  = 2023:2025,
  method = "area",                    # area-weighted mean of overlapping cells
  tz     = "Etc/GMT-12")
```

`method` chooses the spatial sampling:

| `method` | what it does | use when |
|---|---|---|
| `"nearest"` | single closest cell centre | lake well inside one cell |
| `"bilinear"` | 4 cell centres around the centroid | point-like water body (default) |
| `"area_mean"` | unweighted mean of cells overlapping the polygon | lake spans a few cells |
| `"area"` | **area-weighted** mean of overlapping cells | lake spans several cells unevenly |

The bundled Lake Rotorua polygon spans more than one cell, so `"area"` is
the right choice:

```{r inspect-shape, eval = requireNamespace("sf", quietly = TRUE)}
lake_poly <- readRDS(file.path(ex, "rotorua_lake_shape.rds"))
sf::st_crs(lake_poly)$epsg          # 2193 (NZTM); any CRS is accepted
round(as.numeric(sum(sf::st_area(lake_poly))) / 1e6, 1)   # lake area, km^2
sf::st_bbox(sf::st_transform(lake_poly, 4326))            # ~0.11 deg E-W
```

An ~0.11&deg; east--west extent against a ~0.1&deg; grid means the lake
touches two to three cells in each direction --- exactly the case
area-weighting is for.

## Visualising the grid and the sampling

`plot_extract_grid()` reads the same files an extraction would and draws the
point or polygon against the ERA5-Land grid, shading the cells the chosen
`method` samples by their weight. Run it before an extraction to confirm the
geometry sits where you expect --- a shapefile in the wrong projection or a
sign-flipped longitude jumps out on the map and nowhere else.

The package ships only the extracted series, so the chunks below synthesise a
bare ERA5-Land grid (regular 0.1&deg;) over the lake; against real files on
disk the `plot_extract_grid()` calls are identical.

```{r viz-can, eval = TRUE, include = FALSE}
have_viz <- all(vapply(c("sf", "ncdf4", "ggplot2"), requireNamespace,
                       logical(1), quietly = TRUE))
```

```{r grid-fixture, eval = have_viz, include = FALSE}
suppressMessages(library(ggplot2))

## write a single-variable hourly ERA5-Land-style netCDF; `na_cells` blanks
## given (col, row) cells, standing in for ERA5-Land's sea / lake mask
write_era5_land_grid <- function(dir, na_cells = NULL) {
  dir.create(dir, showWarnings = FALSE, recursive = TRUE)
  lon_ax <- seq(175.9, 176.7, by = 0.1)
  lat_ax <- seq(-38.35, -37.75, by = 0.1)
  secs   <- as.numeric(as.POSIXct(c("2024-01-01 00:00", "2024-01-01 01:00"),
                                  tz = "UTC"))
  a <- array(288, c(length(lon_ax), length(lat_ax), length(secs)))
  for (rc in na_cells) a[rc[1], rc[2], ] <- NA
  ld <- ncdf4::ncdim_def("longitude",  "degrees_east",  lon_ax)
  bd <- ncdf4::ncdim_def("latitude",   "degrees_north", lat_ax)
  vd <- ncdf4::ncdim_def("valid_time", "seconds since 1970-01-01 00:00:00", secs)
  vv <- ncdf4::ncvar_def("t2m", "K", list(ld, bd, vd), missval = -9999)
  nc <- ncdf4::nc_create(file.path(dir, "era5-land_2024_01_2m_temperature.nc"),
                         list(vv))
  ncdf4::ncvar_put(nc, vv, a); ncdf4::nc_close(nc)
  dir
}

gdir <- write_era5_land_grid(file.path(tempdir(), "era5_land_grid"))
lake <- readRDS(file.path(ex, "rotorua_lake_shape.rds"))
```

### A point, `"bilinear"`

For `lon` / `lat` (or a small water body) `plot_extract_grid()` shows the four
nodes the bilinear blend uses and their weights.

```{r viz-point, eval = have_viz, fig.height = 3.4, fig.cap = "**Figure 1.** A point (red cross) with `method = \"bilinear\"`: the four surrounding grid cells, weighted by proximity (weights sum to 1). `\"nearest\"` would keep only the single closest of them. The subtitle reports the method, the cell count and (for a one-cell sample) the point-to-centre distance."}
plot_extract_grid(gdir, lon = 176.2717, lat = -38.0790, method = "bilinear")
```

### The four `method` options on the lake polygon

The same Lake Rotorua outline under each `method` from the table above. The
returned object is a `ggplot` carrying the drawn cells as an `sf` on
`attr(., "grid")` (columns `ix`, `iy`, `weight`, `selected`, `land`), so the
four can be pulled apart and faceted:

```{r viz-methods, eval = have_viz, fig.height = 5.4, fig.cap = "**Figure 2.** `nearest` takes the one cell nearest the centroid; `bilinear` blends the four around it; `area_mean` averages every cell the polygon overlaps with equal weight; `area` weights those same cells by the share of the polygon falling in each. Colour is the weight (summing to 1 in every panel). Lake Rotorua covers an uneven 2&times;2 block, so `area` and `area_mean` disagree --- `area` is the honest lake average."}
methods <- c("nearest", "bilinear", "area_mean", "area")

grid_by_method <- do.call(rbind, lapply(methods, function(m) {
  g <- attr(plot_extract_grid(gdir, geom = lake, method = m), "grid")["weight"]
  g$method <- factor(m, levels = methods)
  g
}))
lake_ll <- sf::st_transform(sf::st_geometry(lake), 4326)

ggplot(grid_by_method) +
  geom_sf(aes(fill = weight), colour = "grey80", linewidth = 0.2) +
  geom_sf(data = lake_ll, fill = NA, colour = "#D6201F", linewidth = 0.7) +
  facet_wrap(~ method) +
  scale_fill_viridis_c(option = "C", na.value = "grey92", limits = c(0, NA),
                       name = "weight") +
  coord_sf(expand = FALSE) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 9) +
  theme(axis.text = element_blank(), axis.ticks = element_blank(),
        panel.grid = element_blank())
```

### Cells masked by ERA5-Land

ERA5-Land carries no data over the sea, and over some inland-water pixels.
Masked cells are outlined dashed blue; the extractor drops them and rescales
the surviving weights so they still sum to 1. Blanking the centre cell of the
block:

```{r viz-mask, eval = have_viz, fig.height = 3.6, fig.cap = "**Figure 3.** The centre cell masked (dashed blue) and dropped from `method = \"area\"`; its weight is spread over the three cells that remain. `max_dist_km` / the nearest-valid fallback use the same mask, so a lake sitting entirely on masked pixels still resolves to the closest cell that has data."}
centre <- attr(plot_extract_grid(gdir, geom = lake, method = "nearest"),
               "weights")
gdir_masked <- write_era5_land_grid(
  file.path(tempdir(), "era5_land_masked"),
  na_cells = list(c(centre$ix, centre$iy)))

plot_extract_grid(gdir_masked, geom = lake, method = "area")
```

### Other arguments

* `pad` --- grid cells of context drawn around the geometry (default `3`);
  raise it to see more of the surrounding grid.
* `max_dist_km` --- for a `"nearest"` sample, `plot_extract_grid()` **warns**
  and still draws (with the distance in the subtitle) where
  `extract_era5_hourly_met()` would stop with an error.
* `area_crs` --- projected CRS the `"area"` intersection areas are computed
  in (default `2193`, NZTM 2000); use one suited to your region.
* `engine = "base"` --- draw with `sf::plot()` instead of \pkg{ggplot2} and
  return the grid `sf` invisibly, for a dependency-free check.
* `variables`, `pattern`, `years`, `months` --- passed straight to the file
  matcher, so the preview opens the same files the extraction will.

```{r viz-args, eval = have_viz, fig.height = 3.4, fig.cap = "**Figure 4.** `max_dist_km` with a point ~110 km east of the grid: `plot_extract_grid()` warns, picks the nearest edge cell and draws it so the mistake is visible; `extract_era5_hourly_met(method = \"nearest\")` would abort with the same message."}
plot_extract_grid(gdir, lon = 178.0, lat = -38.0, method = "nearest",
                  max_dist_km = 50, pad = 2)
```

## `extract_era5_lake_met()` --- the polygon wrapper

`extract_era5_lake_met()` is the same call with two conveniences: `method`
defaults to `"area"`, and optional `id` / `name` are stored on the result
as `lake_id` / `lake_name` attributes. The polygon can be an `sf` / `sfc`
object or a path to a vector file (`.gpkg`, `.shp`, `.rds`).

```{r lake-wrapper}
met <- extract_era5_lake_met(
  lake_poly,                     # or "gis/rotorua.gpkg"
  path    = "era5_netcdf",
  years   = 2023:2025,
  name    = "Rotorua",
  outfile = "rotorua_era5_hourly_met.csv")

## batch over a multi-lake layer
lakes <- sf::st_read("gis/lakes.gpkg")
mets  <- lapply(seq_len(nrow(lakes)), function(i)
  extract_era5_lake_met(lakes[i, ], path = "era5_netcdf", years = 2024,
                        id = lakes$id[i], name = lakes$name[i]))
```

The bundled `rotorua_era5_hourly_met.csv.gz` used throughout the other
vignettes is the output of exactly this call for Lake Rotorua.

## See also

* `vignette("download-era5")` --- getting the netCDF in the first place.
* `vignette("met-data-frames")` --- the `MET_*` names, units and time-zone
  convention the extractor output follows.
* `vignette("bias-correction-methods")` --- correcting the extracted series
  against local observations.
* `?plot_extract_grid` --- the pre-extraction grid check shown above.
