---
title: "Meteorology data frames: names, units, time zones"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Meteorology data frames: names, units, time zones}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 90)
```

Every `metscale` function reads and returns the same shape of object: a data
frame with a `Date` column and one column per meteorological variable, named
and unitised by a fixed convention. This vignette covers that convention and
the helpers that bring an arbitrary table into it.

```{r setup}
library(metscale)
ex <- system.file("extdata", package = "metscale")
```

## The `MET_*` scheme

Columns follow the AEME `MET_*` names. `met_vars()` is the full list with
units:

```{r met-vars}
met_vars()
```

A few points that trip people up:

* **Rainfall and snowfall are per-timestep totals**, not rates --- `mm` in
  an hourly frame means mm/hour, `mm` in a daily frame means mm/day. Every
  other variable is an instantaneous value (or an interval mean).
* **Pressure is in pascals.** `MET_prsttn` (station) and `MET_prmslp` (sea
  level) are Pa, not hPa. Vapour pressure `MET_prvapr` is the exception --- hPa.
* **Wind direction is the direction the wind blows *from***, degrees
  clockwise from north, and `MET_wnduvu` / `MET_wnduvv` are the matching
  eastward / northward components.
* **Cloud cover is a 0--1 fraction**, not oktas or percent.

## Matching arbitrary names: `guess_met_vars()`

`guess_met_vars()` maps logger / council / NIWA / ERA5 / CMIP column names
onto `MET_*` by case-insensitive pattern, returning `NA` where nothing fits:

```{r guess}
guess_met_vars(c("AirTemp_C", "WindSpd", "RH_pct", "SolarRad_Wm2",
                 "rain_mm", "t2m", "sfcWind", "junk_column"))
```

## Standardising a whole table: `standardise_met()`

`standardise_met()` renames with `guess_met_vars()` **and** repairs the
common unit mistakes it can detect from the value ranges (Kelvin, hPa,
humidity as a fraction, oktas, km/h, precipitation in metres, radiation in
MJ m-2 day-1). It finds the date column itself and renames it `Date`.

```{r standardise}
raw <- data.frame(
  timestamp   = as.Date("2024-01-01") + 0:4,
  AirTemp_K   = c(291, 289, 292, 288, 293),
  WindSpd_kmh = c(11, 18, 7, 22, 11),
  SolarRad    = c(300, 120, 280, 90, 310),
  RH          = c(0.72, 0.85, 0.61, 0.90, 0.55),
  Rain_mm     = c(0, 12, 0, 25, 0))

std <- standardise_met(raw)
std
```

The messages say what it did; pass `verbose = FALSE` to silence them.
Columns it cannot match are kept unchanged and warned about, so nothing is
lost silently. It does **not** resample or derive missing variables --- that
is `prepare_obs_met()` and `expand_met()`.

## Preparing observations: `prepare_obs_met()`

For a *measured* series destined for bias correction, `prepare_obs_met()`
wraps `standardise_met()` and adds the steps that matter for lining the
observations up with reanalysis:

```{r prepare-obs}
obs <- prepare_obs_met(file.path(ex, "rotorua_buoy_met_aeme_hr.csv.gz"),
                       resample    = "hour",   # bin sub-hourly records to whole hours
                       tz          = "Etc/GMT-12",
                       wind_height = 2,        # 2 m buoy anemometer -> 10 m
                       station     = "rotorua_buoy",
                       verbose     = FALSE)
str(obs, give.attr = FALSE)
attributes(obs)[c("tz", "resolution", "station")]
```

What it adds over `standardise_met()`:

* reads a `.csv` / `.csv.gz` / `.rds` path directly;
* `resample = "hour"` / `"day"` --- means for most variables, **sums** for
  rain and snow;
* `interval = "ending"` (default) labels a sub-hourly record by the hour it
  *ends* in, matching ERA5's accumulated-flux convention so the two join
  cleanly on `Date`;
* `derive = TRUE` fills `MET_wndspd`/`MET_wnddir` &harr; u/v and
  `MET_humrel` &harr; `MET_tmpdew` when one side is present;
* `wind_height` rescales a low anemometer to the 10 m convention
  ([`wind_at_height()`]) --- a buoy at 2 m reads roughly 15--20% low;
* records `tz`, `resolution` and `station` as attributes, which
  `fit_met_bias_correction()` carries into its metadata.

## Time zone

Every `tz` argument defaults to `"UTC"` --- the clock ERA5, ERA5-Land and
CMIP6 are published on --- and resolves the same way everywhere: an
explicit value first, then the `tz` attribute carried on the input frame
(so a zone set once on the extractor output flows through the pipeline),
then UTC. The extractors therefore return the reanalysis with no hidden
shift.

To work in local time, pass the zone your timestamps are in. Prefer a
**fixed** offset such as `"Etc/GMT-12"` (NZST, UTC+12, no daylight saving):
it keeps a gap-free regular sub-daily series, whereas civil time with DST
has a missing hour in spring and a doubled hour in autumn that breaks a
strict hourly index. Use `"Pacific/Auckland"` only if your logger recorded
civil time and you need it back in civil time.

The zone is not cosmetic: `expand_met()`, `calc_cc()` and
`estimate_hourly_swr()` evaluate the solar geometry at the matching UTC
instants, so with a local `tz` the shortwave peak lands at local solar
noon. (The AEME versions of these routines forced the session zone to UTC
and read the wall clock, shifting the solar curve by the UTC offset ---
12 h for New Zealand. `metscale` fixes that; pass the zone your timestamps
are actually in.)

## Daily aggregation: `met_to_daily()`

`met_to_daily()` collapses an hourly (or finer) frame to daily --- means for
everything except rain and snow, which are summed --- and returns `Date` as
class `Date`. `min_frac` drops days that are too incomplete to trust.

```{r to-daily}
era5 <- read.csv(file.path(ex, "rotorua_era5_hourly_met.csv.gz"), check.names = FALSE)
era5$Date <- as.POSIXct(era5$Date, tz = "Etc/GMT-12")

daily <- met_to_daily(era5, tz = "Etc/GMT-12", min_frac = 0.9)
head(daily[, c("Date", "MET_tmpair", "MET_wndspd", "MET_pprain")])
```

## Handing data to CF / CMIP tools: `met_to_cf()` / `cf_to_met()`

`met_to_cf()` renames the `MET_*` columns to the CF / CMIP short names
(`tas`, `pr`, `sfcWind`, `rsds`, ...) and converts each to its CF-canonical
unit; `cf_to_met()` is the exact inverse. The precipitation conversion
needs a timestep (mm per step &harr; `kg m-2 s-1`), inferred from the time
column or passed as `timestep`.

```{r cf-roundtrip}
cf <- met_to_cf(daily[, c("Date", "MET_tmpair", "MET_wndspd", "MET_pprain")])
head(cf)
attr(cf, "cf_units")

back <- cf_to_met(cf)
head(back)
## round-trips to within floating point
max(abs(back$MET_pprain - daily$MET_pprain))
```

`era5_ref_table` is the underlying variable / unit / daily-aggregation map
across ERA5, LakeEnsemblR and AEME:

```{r ref-table}
era5_ref_table
```

## See also

* `vignette("deriving-variables")` --- fill a minimal set out to everything
  a lake model needs with `expand_met()`.
* `vignette("bias-correction-methods")` --- correct a standardised ERA5
  frame against a `prepare_obs_met()` frame.
* `vignette("scenario-workflow")` --- the full bias-correction &rarr;
  delta-change &rarr; disaggregation pipeline.
