---
title: "Climate-scenario meteorology workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Climate-scenario meteorology workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6, fig.height = 3.4, dpi = 96, dev = "png"
)
options(width = 90)
```

Driving a sub-daily lake model under a climate scenario needs three things
done in the right order:

1. **bias-correct** the reanalysis against local observations,
2. apply the climate model's **delta-change** signal to the *corrected*
   baseline (never to the raw reanalysis, and never to the delta itself),
3. **temporally disaggregate** the daily projection back to hourly.

`metscale` covers steps 1 and 3 directly; step 2 is a few lines of arithmetic
shown below. See `?scenario_workflow` for why the correction must come before
the delta.

```{r setup, message = FALSE, warning = FALSE}
library(metscale)
library(ggplot2)

theme_set(
  theme_bw(base_size = 10) +
    theme(panel.grid.minor = element_blank(),
          legend.position  = "top",
          legend.title     = element_blank(),
          strip.background  = element_blank(),
          strip.placement   = "outside")
)

ex   <- system.file("extdata", package = "metscale")
tz   <- "Etc/GMT-12"          # fixed NZST (UTC+12)
lon  <- 176.2717
lat  <- -38.0790
elev <- 279                   # Lake Rotorua surface, m
```

The bundled example data is a deliberately small slice: a three-year hourly
ERA5-Land extract for Lake Rotorua (2023--2025, chosen to overlap the buoy
record), the matching buoy meteorology, and the NIWA CCAM downscaling of one
CMIP6 model (`NorESM2-MM`) cropped to a 3&times;3 grid window and to
2005--2014 (historical) plus 2090--2099 (`ssp245`, `ssp585`). A production run
would use a multi-decade reanalysis record and 20-year climate windows; the
steps are identical.

The hourly ERA5-Land slice was produced by `extract_era5_hourly_met()`
reading monthly ERA5-Land files. The appendix (*Checking the extraction
grid*) shows the one-line `plot_extract_grid()` check --- that the lake
outline sits on the grid cells you expect --- worth running before any
extraction.

## 1. Bias-correct hourly ERA5 against the buoy

The buoy anemometer sits at 2 m; `wind_height = 2` rescales it to the 10 m
convention (neutral log profile) so it is comparable with ERA5 before the
bias correction is fitted.

```{r bias-fit}
obs <- prepare_obs_met(file.path(ex, "rotorua_buoy_met_aeme_hr.csv.gz"),
                       resample = "hour", tz = tz, station = "rotorua_buoy",
                       wind_height = 2, verbose = FALSE)

era5 <- read.csv(file.path(ex, "rotorua_era5_hourly_met.csv.gz"),
                 check.names = FALSE)
era5$Date <- as.POSIXct(era5$Date, tz = tz, format = "%Y-%m-%d %H:%M:%S")
attr(era5, "tz") <- tz; attr(era5, "lat") <- lat; attr(era5, "lon") <- lon

## "scale" (per-month offset, day-of-year loess-smoothed) is the right choice
## for a baseline that will be projected: a fitted CDF (eqm/qdm) extrapolates
## poorly outside its training range. Rainfall is left uncorrected here --- a
## lake buoy under-catches rain, so its totals are not a good target; ERA5's
## own precipitation is carried through.
fit_vars <- c("MET_tmpair", "MET_tmpdew", "MET_wndspd", "MET_radswd",
              "MET_humrel", "MET_prsttn")
bc <- fit_met_bias_correction(era5, obs, vars = fit_vars, method = "scale",
                              by = "doy-loess", verbose = FALSE)
bc$skill[bc$skill$stage %in% c("cv_raw", "cv_corrected"), ]
```

```{r bias-plot, fig.height = 3.6, fig.cap = "**Figure 1.** Leave-one-year-out cross-validated RMSE for each fitted variable, in that variable's own units (free y-axis per panel). Because every year is scored by a model that never saw it, the fall from *raw ERA5* to *bias-corrected* is out-of-sample skill, not in-sample fit. Every variable improves; station pressure improves most, for the reason noted below the figure."}
sk  <- bc$skill
stg <- if (any(sk$stage == "cv_raw")) c("cv_raw", "cv_corrected") else c("raw", "corrected")
sk  <- sk[sk$stage %in% stg, ]
sk$kind <- factor(ifelse(grepl("raw", sk$stage), "raw ERA5", "bias-corrected"),
                  c("raw ERA5", "bias-corrected"))

ggplot(sk, aes(kind, rmse, fill = kind)) +
  geom_col(width = 0.6) +
  facet_wrap(~ variable, scales = "free_y") +
  scale_fill_manual(values = c("raw ERA5" = "grey70",
                               "bias-corrected" = "steelblue")) +
  labs(x = NULL, y = "cross-validated RMSE")
```

The large station-pressure correction is **not** a unit mismatch --- both
series are in pascals. ERA5-Land reports surface pressure at its grid-cell
orography height, which around Lake Rotorua sits roughly 250 m above the
279 m lake surface. That elevation gap appears as a near-constant offset of
about -3 kPa (standard deviation only ~70 Pa, correlation 0.996), which the
monthly `scale` offset removes almost exactly.

Apply the fit to the whole record and expand the dependent variables --- this
corrected hourly series is also the donor for disaggregation in step 3.

```{r bias-apply}
era5_corr <- apply_met_bias_correction(era5, bc, expand = TRUE,
                                       lat = lat, lon = lon, elev = elev,
                                       tz = tz, verbose = FALSE)
```

One winter week shows the correction at work on the hourly series.

```{r correction-week, fig.height = 5, fig.cap = "**Figure 2.** Lake Rotorua, 8--15 July 2024: hourly buoy observations (black) against ERA5-Land before (grey) and after (red) the *scale* correction. The correction lifts the persistently cold ERA5 air temperature onto the buoy, scales up the under-forecast wind, and pulls down the moist bias in relative humidity, while leaving the sub-daily shape of each series intact."}
w0  <- as.POSIXct("2024-07-08", tz = tz)
w1  <- as.POSIXct("2024-07-15", tz = tz)
win <- function(d) d[d$Date >= w0 & d$Date < w1, ]

vs <- c(MET_tmpair = "air temperature (°C)",
        MET_wndspd = "wind speed (m/s)",
        MET_humrel = "relative humidity (%)")

mk <- function(d, src) {
  d <- win(d)
  data.frame(Date     = rep(d$Date, length(vs)),
             value    = unlist(d[names(vs)], use.names = FALSE),
             variable = factor(rep(unname(vs), each = nrow(d)), unname(vs)),
             source   = src)
}
wk <- rbind(mk(obs, "buoy"), mk(era5, "raw ERA5"), mk(era5_corr, "bias-corrected"))
wk$source <- factor(wk$source, c("buoy", "raw ERA5", "bias-corrected"))

ggplot(wk, aes(Date, value, colour = source, linewidth = source)) +
  geom_line() +
  facet_wrap(~ variable, ncol = 1, scales = "free_y", strip.position = "left") +
  scale_colour_manual(values = c(buoy = "black", "raw ERA5" = "grey60",
                                 "bias-corrected" = "firebrick")) +
  scale_linewidth_manual(values = c(buoy = 0.8, "raw ERA5" = 0.4,
                                    "bias-corrected" = 0.4), guide = "none") +
  labs(x = NULL, y = NULL)
```

## 2. Bias-corrected daily baseline

```{r baseline}
baseline <- bias_correct_daily_baseline(era5, bc, lat = lat, lon = lon,
                                        elev = elev, tz = tz, verbose = FALSE)

## keep only fully-populated days
ok <- stats::complete.cases(baseline[c("MET_radswd", "MET_tmpair", "MET_pprain",
                                       "MET_wndspd", "MET_humrel")])
baseline <- baseline[ok, ]
range(baseline$Date)
```

## 3. CMIP6 delta-change on the corrected baseline

Read the model series at the lake with `extract_cmip6_point()` --- it decodes
the 365-day model calendar, samples bilinearly, and returns AEME `MET_*`
names and units.

```{r cmip-read}
cmip <- extract_cmip6_point(file.path(ex, "rotorua_cmip6"),
                            lon = lon, lat = lat,
                            vars = c("MET_tmpair", "MET_pprain", "MET_wndspd",
                                     "MET_radswd", "MET_humrel"),
                            verbose = FALSE)
cmip_by <- split(cmip, cmip$experiment)
vapply(cmip_by, nrow, integer(1))
```

Change factors are formed per calendar month, future window vs model
historical: **additive** for temperature and shortwave (degC, W m^-2^),
**multiplicative** for the bounded/skewed fields (rainfall, wind, humidity).

```{r delta-fns}
REF_YEARS <- 2005:2014
FUT_YEARS <- 2090:2099
KIND <- c(MET_tmpair = "add",   MET_radswd = "add",
          MET_pprain = "ratio", MET_wndspd = "ratio", MET_humrel = "ratio")

monthly_delta <- function(fut, hist) {
  mh <- as.integer(format(hist$Date, "%m")); hy <- as.integer(format(hist$Date, "%Y"))
  mf <- as.integer(format(fut$Date,  "%m")); fy <- as.integer(format(fut$Date,  "%Y"))
  kh <- hy %in% REF_YEARS; kf <- fy %in% FUT_YEARS
  lapply(names(KIND), function(v) {
    a <- tapply(hist[[v]][kh], mh[kh], mean, na.rm = TRUE)
    b <- tapply(fut[[v]][kf],  mf[kf], mean, na.rm = TRUE)
    d <- if (KIND[[v]] == "ratio") b / a else b - a
    stats::setNames(as.numeric(d[as.character(1:12)]), 1:12)
  }) |> stats::setNames(names(KIND))
}

apply_delta <- function(base, delta) {
  mo <- as.integer(format(base$Date, "%m"))
  out <- base
  for (v in names(delta)) {
    f <- delta[[v]][mo]
    out[[v]] <- if (KIND[[v]] == "ratio") out[[v]] * f else out[[v]] + f
  }
  out$MET_humrel <- pmin(pmax(out$MET_humrel, 0), 100)
  for (v in c("MET_radswd", "MET_wndspd", "MET_pprain"))
    out[[v]][out[[v]] < 0] <- 0
  ## regenerate dependents; station pressure carries no CMIP delta
  expand_met(out[, c("Date", "MET_radswd", "MET_tmpair", "MET_pprain",
                     "MET_humrel", "MET_wndspd", "MET_prsttn")],
             lat = lat, lon = lon, elev = elev, tz = tz)
}

deltas <- list(ssp245 = monthly_delta(cmip_by$ssp245, cmip_by$historical),
               ssp585 = monthly_delta(cmip_by$ssp585, cmip_by$historical))
sapply(deltas, function(d) round(mean(d$MET_tmpair), 2))   # annual-mean warming, degC
```

```{r delta-fig, fig.height = 3, fig.cap = "**Figure 3.** Monthly-mean air-temperature change factors added in step 2: 2090--2099 minus the 2005--2014 model baseline, one line per scenario. `ssp585` warms more than `ssp245` in every month. These twelve numbers are what `apply_delta()` adds to the corresponding months of the corrected daily baseline."}
dT <- data.frame(
  month    = rep(1:12, 2),
  delta    = c(deltas$ssp245$MET_tmpair, deltas$ssp585$MET_tmpair),
  scenario = rep(c("ssp245", "ssp585"), each = 12))

ggplot(dT, aes(month, delta, colour = scenario)) +
  geom_line() +
  geom_point(size = 1.8) +
  scale_x_continuous(breaks = 1:12) +
  scale_colour_manual(values = c(ssp245 = "steelblue", ssp585 = "firebrick")) +
  expand_limits(y = 0) +
  labs(x = "month", y = expression(Delta * "T  (" * degree * "C)"))
```

```{r project}
proj <- lapply(deltas, apply_delta, base = baseline)
```

```{r project-fig, fig.height = 3, fig.cap = "**Figure 4.** Effect of the delta-change step over one year: the bias-corrected daily-mean air temperature for 2024 (grey) and the same days after adding the `ssp585` 2090--2099 monthly signal (red). The offset varies by month, so the projected series is warped rather than simply shifted."}
yr <- format(baseline$Date, "%Y") == "2024"
pf <- rbind(
  data.frame(Date = baseline$Date[yr],   tmpair = baseline$MET_tmpair[yr],
             series = "bias-corrected baseline"),
  data.frame(Date = proj$ssp585$Date[yr], tmpair = proj$ssp585$MET_tmpair[yr],
             series = "+ ssp585 delta"))
pf$series <- factor(pf$series, c("bias-corrected baseline", "+ ssp585 delta"))

ggplot(pf, aes(Date, tmpair, colour = series)) +
  geom_line() +
  scale_colour_manual(values = c("bias-corrected baseline" = "grey55",
                                 "+ ssp585 delta" = "firebrick")) +
  labs(x = NULL, y = expression("daily mean air T  (" * degree * "C)"))
```

## 4. Disaggregate the projection to hourly

`disaggregate_met_to_hourly()` borrows the within-day shape of the corrected
ERA5 record (method of fragments) and rebuilds shortwave from solar geometry.
Daily means and rainfall totals are conserved.

```{r disagg}
slice  <- proj$ssp585[format(proj$ssp585$Date, "%Y-%m") == "2024-07", ]
hourly <- disaggregate_met_to_hourly(slice, donor = era5_corr,
                                     method = "fragments", swr = "clearsky",
                                     lat = lat, lon = lon, elev = elev, tz = tz,
                                     seed = 42, expand = TRUE, verbose = FALSE)
nrow(hourly)
```

```{r disagg-fig, fig.height = 3, fig.cap = "**Figure 5.** First three days (1--3 July 2024) of the disaggregated hourly `ssp585` projection. Air temperature carries the diurnal shape of the analogue donor day; shortwave is not borrowed but reconstructed from clear-sky solar geometry, so it starts and ends each day at zero."}
d3 <- hourly[format(hourly$Date, "%Y-%m-%d") <= "2024-07-03", ]
df <- rbind(
  data.frame(Date = d3$Date, value = d3$MET_tmpair, panel = "air temperature (°C)"),
  data.frame(Date = d3$Date, value = d3$MET_radswd, panel = "shortwave (W/m²)"))
df$panel <- factor(df$panel, unique(df$panel))

ggplot(df, aes(Date, value, colour = panel)) +
  geom_line() +
  facet_wrap(~ panel, scales = "free_y", strip.position = "left") +
  scale_colour_manual(values = c("firebrick", "goldenrod"), guide = "none") +
  labs(x = NULL, y = NULL)
```

Zooming out to a full week shows what disaggregation does to each field and
what it leaves alone.

```{r daily-hourly-fig, fig.height = 6, fig.cap = "**Figure 6.** Wettest seven-day window of the projected July 2024 month: the daily projection (grey step) and the hourly series disaggregated from it (colour). Hourly air temperature and wind vary about the daily mean; hourly rainfall is redistributed within each day while that day's total (labelled, mm/day) is left unchanged."}
## pick the wettest 7-day window of the month so the rain panel is not empty
roll7 <- sapply(seq_len(nrow(slice) - 6),
                function(i) sum(slice$MET_pprain[i:(i + 6)]))
w0d <- slice$Date[which.max(roll7)]
h7  <- hourly[as.Date(hourly$Date) >= w0d & as.Date(hourly$Date) < w0d + 7, ]
day_of <- function(v) slice[[v]][match(as.Date(h7$Date), as.Date(slice$Date))]

lab <- c(MET_tmpair = "air temperature (°C)",
         MET_wndspd = "wind speed (m/s)",
         MET_pprain = "rainfall (mm/hour)")

hourly_df <- rbind(
  data.frame(Date = h7$Date, value = h7$MET_tmpair, panel = lab[["MET_tmpair"]]),
  data.frame(Date = h7$Date, value = h7$MET_wndspd, panel = lab[["MET_wndspd"]]),
  data.frame(Date = h7$Date, value = h7$MET_pprain, panel = lab[["MET_pprain"]]))
daily_df <- rbind(
  data.frame(Date = h7$Date, value = day_of("MET_tmpair"), panel = lab[["MET_tmpair"]]),
  data.frame(Date = h7$Date, value = day_of("MET_wndspd"), panel = lab[["MET_wndspd"]]))
hourly_df$panel <- factor(hourly_df$panel, lab)
daily_df$panel  <- factor(daily_df$panel,  lab)

## per-day rain totals, placed near the top of the rainfall panel
dd  <- unique(as.Date(h7$Date))
tot <- slice$MET_pprain[match(dd, as.Date(slice$Date))]
rain_lab <- data.frame(
  Date  = as.POSIXct(paste(dd, "02:00:00"), tz = tz),
  value = max(h7$MET_pprain, na.rm = TRUE) * 0.95,
  panel = factor(lab[["MET_pprain"]], lab),
  text  = sprintf("%.1f mm/day", tot))
rain_lab <- rain_lab[tot > 0.05, ]

ggplot(hourly_df, aes(Date, value)) +
  geom_step(data = daily_df, colour = "grey45", linewidth = 0.9,
            direction = "hv") +
  geom_line(data = subset(hourly_df, panel != lab[["MET_pprain"]]),
            aes(colour = panel), linewidth = 0.4) +
  geom_col(data = subset(hourly_df, panel == lab[["MET_pprain"]]),
           fill = "seagreen", width = 3000) +
  geom_text(data = rain_lab, aes(label = text), size = 2.5, hjust = 0,
            colour = "grey35", vjust = 1) +
  facet_wrap(~ panel, ncol = 1, scales = "free_y", strip.position = "left") +
  scale_x_datetime(expand = ggplot2::expansion(mult = c(0.01, 0.03))) +
  scale_colour_manual(values = c("firebrick", "steelblue"), guide = "none") +
  labs(x = NULL, y = NULL)
```

Conservation check --- the hourly series re-aggregated to daily reproduces
the projected daily input:

```{r conservation}
back <- met_to_daily(hourly, tz = tz, min_frac = 1)
m <- match(as.Date(back$Date), as.Date(slice$Date))
data.frame(
  variable = c("MET_tmpair", "MET_humrel", "MET_pprain"),
  max_abs_daily_error = sapply(c("MET_tmpair", "MET_humrel", "MET_pprain"),
                               function(v) max(abs(back[[v]] - slice[[v]][m])))
)
```

## Without local observations

Only step 1 needs observations. The delta-change (step 2) touches nothing but
the climate model runs, and the disaggregator (step 3) only needs an hourly
*donor* at the location for the within-day shape --- raw ERA5-Land is exactly
that. So with no buoy you can still run the pipeline: use raw ERA5 aggregated
to daily as the baseline, apply the same `deltas`, and disaggregate with raw
ERA5 as the donor.

```{r obs-free}
baseline_raw <- met_to_daily(era5, tz = tz)
ok <- stats::complete.cases(baseline_raw[c("MET_radswd", "MET_tmpair", "MET_pprain",
                                           "MET_wndspd", "MET_humrel")])
baseline_raw <- baseline_raw[ok, ]

proj_raw   <- lapply(deltas, apply_delta, base = baseline_raw)
slice_raw  <- proj_raw$ssp585[format(proj_raw$ssp585$Date, "%Y-%m") == "2024-07", ]
hourly_raw <- disaggregate_met_to_hourly(slice_raw, donor = era5,
                                         method = "fragments", swr = "clearsky",
                                         lat = lat, lon = lon, elev = elev, tz = tz,
                                         seed = 42, expand = TRUE, verbose = FALSE)
```

The obs-free run is missing only the step-1 correction, so the difference
between the two hourly series is essentially the ERA5-Land local bias --- here
the roughly 3 &deg;C winter cold bias in air temperature seen in Figure 1.

```{r obs-free-fig, fig.height = 3, fig.cap = "**Figure 7.** The July 2024 `ssp585` projection disaggregated to hourly two ways: with the buoy bias correction (red, as in Figure 5) and straight from raw ERA5-Land with no observations (grey). Delta-change and disaggregation are identical between the two; the near-constant gap is the ERA5-Land air-temperature bias that step 1 would remove."}
cmp <- rbind(
  data.frame(Date = hourly$Date,     value = hourly$MET_tmpair,
             forcing = "obs-based"),
  data.frame(Date = hourly_raw$Date, value = hourly_raw$MET_tmpair,
             forcing = "obs-free (raw ERA5)"))
cmp <- cmp[format(cmp$Date, "%Y-%m-%d") <= "2024-07-03", ]

ggplot(cmp, aes(Date, value, colour = forcing)) +
  geom_line() +
  scale_colour_manual(values = c("obs-based" = "firebrick",
                                 "obs-free (raw ERA5)" = "grey45")) +
  labs(x = NULL, y = "air temperature (°C)")
```

A middle option, when you have observations somewhere useful but not at the
site itself --- a nearby land station, a gap-filled station product --- is to
fit step 1 against that series instead of an on-site sensor; the code is
unchanged.

## Caveats

Both the local correction and the delta-change assume **stationarity** --- a
transfer function fitted on a few years of observations is asserted to hold
under a future climate. Use a long reanalysis record and 20-year climate
windows, prefer `method = "scale"` over quantile mapping for the projected
baseline, and treat the scenario spread (here `ssp245` vs `ssp585`) as the
lower bound on uncertainty. See `?scenario_workflow`. For projections aimed
at storms, droughts or heatwaves rather than the mean climate, see
`vignette("climate-extremes")`.

### Why delta-change and not the model output directly

Steps 2--3 warp an observation-anchored baseline with a monthly change
signal rather than driving the lake model with the CMIP6 series itself. A
climate model is reliable for the *change* between two of its own climates,
not for the *absolute* meteorology at a point:

* **Raw model output is biased in absolute terms** --- mean offset, a
  distorted seasonal cycle, drizzle-heavy rainfall, the wrong wind
  climatology. A sub-daily lake model responds to absolute forcing levels,
  so those errors would land straight in stratification, ice and
  evaporation. Differencing two model climates cancels most of the shared
  bias and keeps only the part worth trusting.
* **The corrected baseline carries real weather** --- genuine day-to-day
  persistence, event structure, inter-variable covariance and a real
  diurnal cycle, all on actual calendar dates. The future model window is a
  different draw of internal variability on a 365-day calendar that matches
  no real sequence. Warping the baseline shifts its statistics while
  keeping a physically consistent series.
* **Resolution** --- the lake model needs hourly forcing; the model series
  here is daily on a coarse grid with no credible sub-daily information.
  Step 4 borrows the within-day shape from the *corrected* reanalysis
  instead of inventing it.
* **Order of operations** --- the reanalysis-to-observation correction is an
  *absolute-level* adjustment, the delta a *difference or ratio*. Adding a
  level correction onto a change factor is undefined: correct the baseline
  first, shift second (`?scenario_workflow`).

### Extremes and storm events

Delta-change rescales the *observed* sequence --- every projected day is a
real past day plus that month's mean shift, or times its ratio. It cannot
move a storm onto a different date, change how many wet or calm days a
month holds, or alter day-to-day variance beyond the single multiplicative
factor on rainfall and wind. The projected return period of a given daily
rainfall or wind total is close to the observed one, only nudged; and the
step-4 disaggregator conserves each daily total while borrowing an observed
within-day profile, so it does not sharpen sub-daily peaks either.

If extremes are the research question --- flood-driving rainfall, storm
mixing, heatwave stratification --- the monthly-mean delta is too blunt.
`vignette("climate-extremes")` works through the alternatives with the same
bundled data: per-quantile change factors, driving from bias-corrected
daily model output, and single-event storyline perturbation, plus how to
diagnose the resulting forcing and feed it to a lake model.

## Appendix --- checking the extraction grid

The hourly ERA5-Land series bias-corrected in step 1
(`rotorua_era5_hourly_met.csv.gz`) was pulled from monthly ERA5-Land files
with `extract_era5_hourly_met()` --- for a lake, its wrapper
`extract_era5_lake_met()`, which averages every grid cell the outline
overlaps. Before running that it is worth a look at *which* cells the
outline actually lands on: a shapefile in the wrong projection, a
longitude on the wrong side of the antimeridian, or a lake smaller than
one 0.1&deg; cell all show up instantly on a map and silently otherwise.

`plot_extract_grid()` reads the same files the extractor would, draws the
point or polygon against the grid, and shades the cells the chosen `method`
would sample --- one cell for `"nearest"`, up to four for `"bilinear"`,
every overlapping cell for `"area"` / `"area_mean"`, each weighted by its
share of the intersection. Cells masked out by ERA5-Land (sea, or the lake
pixels themselves) are outlined in dashed blue and dropped, their weight
spread over the rest.

```{r grid-fixture, include = FALSE}
## ERA5-Land is a regular 0.1-degree grid. The package ships only the
## extracted series, not the source netCDF, so synthesise a bare grid over
## the lake for this illustration; against real ERA5-Land files on disk the
## call below is exactly the same.
have_grid <- requireNamespace("ncdf4", quietly = TRUE) &&
             requireNamespace("sf", quietly = TRUE)
if (have_grid) {
  grid_dir <- file.path(tempdir(), "era5_land_grid")
  dir.create(grid_dir, showWarnings = FALSE)
  lon_ax <- seq(175.9, 176.7, by = 0.1)
  lat_ax <- seq(-38.35, -37.75, by = 0.1)
  secs   <- as.numeric(as.POSIXct(c("2024-01-01 00:00", "2024-01-01 01:00"),
                                  tz = "UTC"))
  ld <- ncdf4::ncdim_def("longitude",  "degrees_east",  lon_ax)
  bd <- ncdf4::ncdim_def("latitude",   "degrees_north", lat_ax)
  vd <- ncdf4::ncdim_def("valid_time", "seconds since 1970-01-01 00:00:00", secs)
  vv <- ncdf4::ncvar_def("t2m", "K", list(ld, bd, vd), missval = -9999)
  nc <- ncdf4::nc_create(
    file.path(grid_dir, "era5-land_2024_01_2m_temperature.nc"), list(vv))
  ncdf4::ncvar_put(nc, vv,
                   array(288, c(length(lon_ax), length(lat_ax), length(secs))))
  ncdf4::nc_close(nc)
}
```

```{r grid-check, eval = have_grid, fig.height = 4, fig.cap = "**Figure 8.** The Lake Rotorua outline (red, centroid crossed) on the ERA5-Land 0.1&deg; grid. `method = \"area\"` samples the four cells the polygon overlaps and weights each by its share of the intersection (legend) --- so the extracted lake series is that area-weighted average. Grey dots are cell centres. Here the outline is correctly placed and spans a sensible 2&times;2 block; a projection or coordinate mistake would put it somewhere else entirely."}
lake <- readRDS(file.path(ex, "rotorua_lake_shape.rds"))

plot_extract_grid(grid_dir, geom = lake, method = "area")
```

`plot_extract_grid()` returns a `ggplot`, so it takes further `+` layers,
and it carries the drawn cells as an `sf` on `attr(., "grid")` (columns
`ix`, `iy`, `weight`, `selected`, `land`) --- hand that to `mapview` or
`tmap` for an interactive check against a basemap without either becoming a
package dependency. Pass `engine = "base"` for a dependency-free
`sf::plot()` version.
